\onehalfspacing

\pagenumbering{gobble}

\vspace*{\fill}
\begin{center}
  \Large{\textbf{Internship report}}\\
  \vspace*{1\baselineskip}
  Vo Van Nghia\\
  \vfill
  \vspace*{\fill}
  \Large{\textbf{Date}}\\
  29 Sep, 2022
\end{center}

\newpage

\newpage
\pagenumbering{roman}
\renewcommand{\contentsname}{Table of contents}
\tableofcontents

\newpage

\begin{center}
  \underline{\textbf{ACKNOWLEDGEMENT}}
\end{center}
I would like to thank my supervisors: Urtzi AYESTA, Maaike VERLOOP and Matthieu JONCKHEERE for giving me an opportunity to work on this internship as well as your guidance and patience during the whole internship !

\newpage

\begin{center}
  \underline{\textbf{Abstract}}
\end{center}
Markov decision processes (MDPs) and their model free counterpart in reinforcement learning (RL) have known a large success in the last two decades. Although research in these two areas has been taking place for more than fifty years, the field gained momentum only recently following the advent of powerful hardware and algorithms with which suprahuman performance were obtained in games like Chess or Go. However, these impressive successes often rely on quite exceptional hardware possibilities and cannot be applied in many usual contexts, where, for instance, the volume of data available or the amount of computing power is more restricted. To define the next generation of more democratic and widely applicable algorithms, such methods still need to deal with very demanding exploration issues as soon as the state/action spaces are not small.

The internship aims to improve the efficiency of learning algorithms by leveraging the structure of the underlying problem and focus mainly on model free approach. The final report is divided into 8 parts. After a first introduction about the laboratory, we then present two situations where we study more carefully. Next, a short recap about MDPs and reinforcement learning, as well as discussions about three learning methods used in this report are presented. Finally, we compare the result of each method and have a final conclusion.

\newpage

\pagenumbering{arabic}
